Commutative character sheaves and geometric types for supercuspidal representations

### 2017-08-30 Geometric types.tex
Cleaning up while working with David

In this paper we follow the suggestion above by showing that certain Yu type data are geometrizable, in the following sense.
A Yu type datum determines a sequence of representations $\orho_i$ of compact $p$-adic groups $\oK^i$, for $i=0, \ldots, d$, such that $(\oK^d, \rho_d)$ is a type for a supercuspidal representation of a $p$-adic group.
Let $R$ be the ring of integers of a local field with finite residue field $k$.
The main result of \cite{Yu:models} shows how to find, for each $i=0, \ldots, d$, a smooth group scheme $\underline{G}^i$ over the ring $R$ with $\underline{G}^i(R)=\,^\circ K^i $.
Under certain assumptions on the Yu type datum, we show how each representation $\orho_i$ can be replaced by a pair $(\underline{G}^i, \cs{F}^i)$, where $\cs{F}^i$ is a formal ${\bar\QQ}$-linear conbination of conjugation-equivariant sheaf complexes on the Greenberg transform $G^i$ of $\underline{G}^i$.
%
Writing $\trFrob{\cs{F}^i}$ for the function on $G^i(k) = \underline{G}^i(R) = \,^\circ K^i$ obtained by evaluating the trace of the action of Frobenius on the rational virtual complex $\cs{F}^i$, we show in Theorem~\ref{thm:geotypes} that 
\begin{equation}\label{eqn:intro1}
\trFrob{\cs{F}^i} = \trace(\,^\circ\rho_i).
\end{equation}
By this theorem, then, we obtain geometric avatars for each type in a Yu datum:
\[
\begin{tikzcd}
(\oK^i,\orho_i) \arrow[bend left, dashed]{rrr}{\text{geometrization}} &&& \arrow[bend left, dashed]{lll}{\text{trace of Frob}}
(\underline{G}^i, \cs{F}^i).
\end{tikzcd}
\]
We refer to the pair $(\underline{G}^d, \cs{F}^d)$ as a \emph{geometric type}.

@@

\begin{corollary} 
If $G$ is a smooth group scheme and $(\cs{L},\mu) \in \CS(G)$, then
the restriction of $\cs{L}$ to $G\der$ is trivial if and only if $(\cs{L},\mu) \cong q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$, for some $(\cs{L}\ab,\mu\ab) \in \CS(G\ab)$.
\end{corollary}

@@

Conversely, suppose $(\cs{L},\mu)\in \CS(G)$ and $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$.
Then 
\[
j\der^*(\cs{L},\mu) \iso j\der^*q^*(\cs{L}\ab,\mu\ab)
\]
 in $\CS(G\der)$.
Since $q\circ j\der = 1$, it follows that $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$.
\end{proof}

@@

For connected reductive groups $G$ over $k$, Lusztig's character sheaves are simple perverse sheaves $\IC(C,\mathcal{L})$ on $G$ that come equipped with an isomorphism $a^*\, \IC(C,\mathcal{L}) \to p_2^*\, \IC(C,\mathcal{L})$ in the triangulated category $D^b_c(G\times G;\EE)$.
Consequently, when $\IC(C,\mathcal{L})$ is also a Weil sheaf complex, its trace of Frobenius function $\trFrob{\IC(C,\cs{L})} : G(k)\to \EE$ is a class function, called the characteristic function of $\IC(C,\cs{L})$.
By \cite{lusztig:86a}*{Cor. 25.7}, Weil character sheaves on $G$ determine a basis for the vector space of class functions on $G(k)$; in particular, any character of any representation of $G(k)$ can be expressed as a ${\bar \QQ}$-linear combination of characteristic functions of Weil character sheaves on $G$.
Let us use the notation $\mathsf{D}^b_{c,G}(G;\EE)$ for the category of conjugation-equivariant objects in $\mathsf{D}^b_{c}(G;\EE)$, as defined in \cite{bernstein-lunts:equivariant} for example, and $\mathsf{KD}^b_{c,G}(G;\EE)$ for the Grothendieck group of $\mathsf{D}^b_{c,G}(G;\EE)$; we set $\mathsf{K}_{\bar\QQ}\textsc{D}^b_{c,G}(G;\EE)\ceq \mathsf{KD}^b_{c,G}(G;\EE)\times_{\ZZ} {\bar\QQ}$.
Thus, for every representation $\rho$ of $G(k)$ there is some $\cs{F}\in \mathsf{K}_{\bar\QQ}\textsc{D}^b_{c,G}(G;\EE)$ such that $\trace\rho = \trFrob{\cs{F}}$.
This result extends to reductive groups $G$ over $k$ for which the geometric component group for $G$ is cyclic \cite{lusztig:disconnected1}.

@@

\begin{theorem}\label{thm:geotypes}
Let $(\oK^i, \orho^0, \varphi^i,d)$ be a Yu type datum that satisfies Hypotheses~\ref{H0} and \ref{H1}, and let $\orho_i$ be the representation of $\oK^i$ constructed from it in Section~\ref{ssec:review}.  Then for each $i=0,\dots, d$ there is an element $\cs{F}_i \in \mathcal{K}_{\bar\QQ} D_{G^i}(G^i)$ such that
\[
\trFrob{\cs{F}_i} = \trace(\orho_i).
\]
\end{theorem}

@@

By \cite{lusztig:disconnected1}, there is a rational virtual Weil character sheaf $A = ({\bar A},\phi)$ on $(\underline{G}^0)_\Fq^\reductive$ such that 
%${\bar A}$ is a rational virtual character sheaf on $(\underline{G}^0)_{\bFq}^\reductive$ and
\[
\trFrob{A} = \trace \varrho_0^\reductive.
\]
This uses Hypothesis~\ref{H0}.
%
Let $A^0$ be the conjugation-equivariant Weil sheaf on $(\underline{G}^0)_\Fq$ obtained by pullback along the quotient $(\underline{G}^0)_\Fq \to (\underline{G}^0)_\Fq^\reductive$.
Then 
\[
\trFrob{A^0} = \trace \varrho_0.
\]
The special fibre $(\underline{G}^0)_\Fq$ of the smooth group scheme $\underline{G}^0$ is itself a smooth group scheme, and may be identified with the Greenberg transform $Q^0 = \Gr^{R}_0(\underline{G}^0)$ \cite{cunningham-roe:13a}*{\S 4.3}. 
With $r\in \NN$ as above, let ${A}_r^0$ be the equivariant Weil sheaf on the algebraic group $G_r^i$ obtained by pull-back from $A^0$ along the affine morphism $G_r^i \to Q^0$.
Factor
\begin{equation}\label{eqn:tracerho}
\begin{tikzcd}
G^0(\Fq) \arrow{rr}{\trace(\orho^0)} \arrow{rd} && \EE \\
& G_r^0(\Fq) \arrow{ru}[swap]{\trace(\orho^0)_r} 
\end{tikzcd}
\end{equation}
Observe that $\trace(\orho^0)_r$ may be recovered from ${A}_m^0$:
\[
\trFrob{{A}^0_r} = \trace(\orho^0)_r
\]

@@

\bib{KR:l'adic}{book}{
   author={Kiehl, Reinhardt},
   author={Weissauer, Rainer},
   title={Weil conjectures, perverse sheaves and $l$'adic Fourier transform},
   series={Ergebnisse der Mathematik und ihrer Grenzgebiete. 3. Folge. A
   Series of Modern Surveys in Mathematics [Results in Mathematics and
   Related Areas. 3rd Series. A Series of Modern Surveys in Mathematics]},
   volume={42},
   publisher={Springer-Verlag, Berlin},
   date={2001},
   pages={xii+375},
%   isbn={3-540-41457-6},
%   review={\MR{1855066}},
%   doi={10.1007/978-3-662-04576-3},
}

@@

\bib{lusztig:86a}{article}{
	Author = {George Lusztig},
	Date-Added = {2017-06-28 19:26:19 +0000},
	Date-Modified = {2017-06-28 19:28:23 +0000},
	Journal = {Advances in Math.},
	Keywords = {qcsheaves},
	Number = {2},
	Pages = {{103--155}},
	Title = {{Character sheaves V}},
	Volume = {61},
	Year = {1986}}

>>>>

In this paper we follow the suggestion above by showing that certain Yu type data are geometrizable, in the following sense.
A Yu type datum determines a sequence of representations $\orho_i$ of compact $p$-adic groups $\oK^i$, for $i=0, \ldots, d$, such that $(\oK^d, \rho_d)$ is a type for a supercuspidal representation of a $p$-adic group.
Let $R$ be the ring of integers of a local field with finite residue field $k$.
The main result of \cite{Yu:models} shows how to find, for each $i=0, \ldots, d$, a smooth group scheme $\underline{G}^i$ over the ring $R$ with $\underline{G}^i(R)=\,^\circ K^i $.
Under certain assumptions on the Yu type datum, we show how each representation $\orho_i$ can be replaced by a pair $(\underline{G}^i, \cs{F}^i)$, where $\cs{F}^i$ is a formal ${\bar\QQ}$-linear conbination of conjugation-equivariant sheaf complexes on the Greenberg transform $G^i$ of $\underline{G}^i$.
%
Writing $\trFrob{\cs{F}^i}$ for the function on $G^i(k) = \underline{G}^i(R) = \,^\circ K^i$ obtained by evaluating the trace of the action of Frobenius on $\cs{F}^i$, we show in Theorem~\ref{thm:geotypes} that 
\begin{equation}\label{eqn:intro1}
\trFrob{\cs{F}^i} = \trace(\,^\circ\rho_i).
\end{equation}
By this theorem, then, we obtain geometric avatars for each type in a Yu datum:
\[
\begin{tikzcd}
(\oK^i,\orho_i) \arrow[bend left, dashed]{rrr}{\text{geometrization}} &&& \arrow[bend left, dashed]{lll}{\text{trace of Frob}}
(\underline{G}^i, \cs{F}^i).
\end{tikzcd}
\]
We refer to the pair $(\underline{G}^d, \cs{F}^d)$ as a \emph{geometric type}.

@@

\begin{corollary} 
If $G$ is a smooth group scheme and $(\cs{L},\mu) \in \CS(G)$, then
the restriction of $(\cs{L},\mu)$ to $G\der$ is trivial if and only if $(\cs{L},\mu) \cong q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$, for some $(\cs{L}\ab,\mu\ab) \in \CS(G\ab)$.
\end{corollary}

@@

Conversely, suppose $(\cs{L},\mu)\in \CS(G)$ and $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$.
Then 
\[
j\der^*(\cs{L},\mu) \iso j\der^*q^*(\cs{L}\ab,\mu\ab)
\]
 in $\CS(G\der)$.
Since $q\circ j\der = 1$, it follows that the restriction of $(\cs{L},\mu)$ to $G\der$ is trivial.
\end{proof}

@@

For connected reductive groups $G$ over $k$, Lusztig's character sheaves are simple perverse sheaves $\IC(C,\mathcal{L})$ on $G$ that come equipped with an isomorphism $a^*\, \IC(C,\mathcal{L}) \to p_2^*\, \IC(C,\mathcal{L})$ in the triangulated category $D^b_c(G\times G;\EE)$.
Consequently, when $\IC(C,\mathcal{L})$ is also a Weil sheaf complex, its trace of Frobenius function $\trFrob{\IC(C,\cs{L})} : G(k)\to \EE$ is a class function, called the characteristic function of $\IC(C,\cs{L})$.
By \cite{lusztig:86a}*{Cor. 25.7}, Weil character sheaves on $G$ determine a basis for the vector space of class functions on $G(k)$; in particular, any character of any representation of $G(k)$ can be expressed as a ${\bar \QQ}$-linear combination of characteristic functions of Weil character sheaves on $G$.
Let us use the notation $\mathsf{D}^b_{c,G}(G;\EE)$ for the category of conjugation-equivariant objects in $\mathsf{D}^b_{c}(G;\EE)$, as defined in \cite{bernstein-lunts:equivariant} for example, and $\mathsf{KD}^b_{c,G}(G;\EE)$ for the Grothendieck group of $\mathsf{D}^b_{c,G}(G;\EE)$; we set $\mathsf{K}_{\bar\QQ}\textsc{D}^b_{c,G}(G;\EE)\ceq \mathsf{KD}^b_{c,G}(G;\EE)\times_{\ZZ} {\bar\QQ}$.
Thus, for every representation $\rho$ of $G(k)$ there is some $\cs{F}\in \mathsf{K}_{\bar\QQ}\textsc{D}^b_{c,G}(G;\EE)$ such that $\trace\rho = \trFrob{\cs{F}}$.
This result is extended in the papers beginning with \cite{lusztig:disconnected1} to reductive groups $G$ over $k$ for which the geometric component group for $G$ is cyclic .

@@

\begin{theorem}\label{thm:geotypes}
Let $(\oK^i, \orho^0, \varphi^i,d)$ be a Yu type datum that satisfies Hypotheses~\ref{H0} and \ref{H1}, and let $\orho_i$ be the representation of $\oK^i$ constructed from it in Section~\ref{ssec:review}.  Then for each $i=0,\dots, d$ there is an element $\cs{F}_i \in \mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,G^i}(G^i;\EE)$ such that
\[
\trFrob{\cs{F}_i} = \trace(\orho_i).
\]
\end{theorem}

@@

By \cite{lusztig:disconnected1}, there is $A \in \mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,(\underline{G}^0)_\Fq\red}((\underline{G}^0)_\Fq\red;\EE)$ such that 
%${\bar A}$ is a rational virtual character sheaf on $(\underline{G}^0)_{\bFq}^\reductive$ and
\[
\trFrob{A} = \trace \varrho_0^\reductive.
\]
This uses Hypothesis~\ref{H0}.
%
Let $A^0\in \mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,(\underline{G}^0)_\Fq}((\underline{G}^0)_\Fq;\EE)$ be obtained by pullback along the quotient $(\underline{G}^0)_\Fq \to (\underline{G}^0)_\Fq^\reductive$.
Then 
\[
\trFrob{A^0} = \trace \varrho_0.
\]
The special fibre $(\underline{G}^0)_\Fq$ of the smooth group scheme $\underline{G}^0$ is itself a smooth group scheme, and may be identified with the Greenberg transform $Q^0 = \Gr^{R}_0(\underline{G}^0)$ \cite{cunningham-roe:13a}*{\S 4.3}. 
With $r\in \NN$ as above, let ${A}_r^0$ be the equivariant Weil sheaf on the algebraic group $G_r^i$ obtained by pull-back from $A^0$ along the affine morphism $G_r^i \to Q^0$.
Factor
\begin{equation}\label{eqn:tracerho}
\begin{tikzcd}
G^0(\Fq) \arrow{rr}{\trace(\orho^0)} \arrow{rd} && \EE \\
& G_r^0(\Fq) \arrow{ru}[swap]{\trace(\orho^0)_r} 
\end{tikzcd}
\end{equation}
Observe that $\trace(\orho^0)_r$ may be recovered from ${A}_m^0$:
\[
\trFrob{{A}^0_r} = \trace(\orho^0)_r
\]

@@

\iffalse
\bib{KR:l'adic}{book}{
   author={Kiehl, Reinhardt},
   author={Weissauer, Rainer},
   title={Weil conjectures, perverse sheaves and $l$'adic Fourier transform},
   series={Ergebnisse der Mathematik und ihrer Grenzgebiete. 3. Folge. A
   Series of Modern Surveys in Mathematics [Results in Mathematics and
   Related Areas. 3rd Series. A Series of Modern Surveys in Mathematics]},
   volume={42},
   publisher={Springer-Verlag, Berlin},
   date={2001},
   pages={xii+375},
%   isbn={3-540-41457-6},
%   review={\MR{1855066}},
%   doi={10.1007/978-3-662-04576-3},
}
\fi

@@

\bib{lusztig:86a}{article}{
	Author = {George Lusztig},
%	Date-Added = {2017-06-28 19:26:19 +0000},
%	Date-Modified = {2017-06-28 19:28:23 +0000},
	Journal = {Advances in Math.},
%	Keywords = {qcsheaves},
	Number = {2},
	Pages = {{103--155}},
	Title = {{Character sheaves V}},
	Volume = {61},
	Year = {1986}}
